---
title: "Ant microbes R Project"
author: "Ilia Popov"
date: "2024-09-01"
output:
  html_document:
    latex_engine: xelatex
    df_print: default
    highlight: zenburn
    toc: yes
    toc_depth: 3
  pdf_document:
    toc: yes
    toc_depth: '3'
bibliography: packages.bib
mainfont: NanumGothic
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
if (!require("pacman")) install.packages("pacman")

pacman::p_load(ggplot2, dplyr, stats, psych, readr, tidyr, ggpubr, purrr, rstatix)
```

```{r}
main_dir <- dirname(rstudioapi::getSourceEditorContext()$path) 
setwd(main_dir)
```

# **Part 1 - pH of the ant goitre**

Questions to be answered:<br>

1. Does the acidity of the goitre depend on feeding?<br>
2. At what point in time is the pH lowest?<br>
3. Does the acidity differ between ants from different colonies?<br>

```{r}
ant_data_pH <- read.csv("data/Supplementary_Data_Fig.1a.txt", sep="\t")
```

## **EDA**

```{r}
summary(ant_data_pH)
```


```{r}
describe(ant_data_pH)
```

```{r}
str(ant_data_pH)
```

The dataset contains 231 entries for the _Camponotus floridanus_ ants, detailing their `colony`, `worker` ID, `time` of pH measurement, and the `pH` values themselves. Here are a few observations and next steps:<br>

**Species Column**: All entries belong to the species _c.floridanus_, so we don't need to account for different species in our analysis.<br>
**Time and pH**: The dataset records the pH of the goitre at different time points after feeding (0+4h, 0+24h, 0+48h, 48+4h). We'll compare these to see if feeding time influences pH.<br>
**pH Values**: pH ranges from 1.5 to 5.0. The mean pH is approximately 3.35, suggesting a generally acidic environment, which is expected given that formic acid is a primary component of the gland secretion mentioned.<br>

## **Visualisation**

Let's start with visualizing the pH distribution overall and at each specified time to see if there's an apparent trend or shift related to feeding times. We'll use histograms and box plots for these visualizations.

```{r, warning=FALSE}
ggplot(ant_data_pH, aes(x = ph, fill = time)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.1, alpha = 0.6, position = "identity") +
  geom_density(alpha = 0.5) +
  facet_wrap(~time) +
  labs(title = "Density and Distribution of pH over Time")
```

```{r}
level_order <- c('0+4h', '0+24h', '0+48h', '48+4h')

ggplot(ant_data_pH, aes(x = time, y = ph)) +
  scale_x_discrete(limits = level_order)+
  geom_boxplot() +
  labs(title = "Boxplot of pH Values by Time")
```

### **Visualisation results**
**Density Plot**: Shows variations in the pH distribution at each time point. We can observe shifts in pH concentration and spread, indicating possible changes due to feeding.<br>
**Boxplot**: Highlights median values and spread of pH at each time point. Notably, the interquartile range and median values vary, suggesting significant changes in pH levels related to the time post-feeding.<br>
**Data distribution**: Non-normal.<br>

## **Normality test**

Let's use `shapiro.test` to check the distribution of pH values.

```{r}
shapiro.test(ant_data_pH$ph)
```

### **Normality test results**

The data is distributed non-normal

## **Statistical Analysis**

**Kruskal-Wallis rank sum test for Time Dependency**: We'll conduct an Kruskal-Wallis rank sum test to test if there are statistically significant differences in pH levels at different times.<br>
**Kruskal-Wallis rank sum test for Colony Differences**: Another Kruskal-Wallis rank sum test will help determine if pH varies significantly across different ant colonies.<br>
**Reasons of choice**: Non-normal distributed data. Data with independent observations.<br>

```{r}
kruskal.test(ph ~ time, data = ant_data_pH)
```

```{r}
kruskal.test(ph ~ colony, data = ant_data_pH)
```

### **Statistical Analysis Results**

**Kruskal-Wallis rank sum test for Time Dependency**:
The results show _p_-value close to zero ( _p_ < 0.001 ). This strongly suggests that the pH levels vary significantly across different times, confirming the impact of feeding time on the goitre's acidity.<br>
**Kruskal-Wallis rank sum test for Colony Differences**:
The Kruskal-Wallis comparing different colonies shows a non-significant result ( _p_ = 0.983 ). This implies that the differences in pH levels between colonies are not statistically significant, suggesting that colony does not influence the acidity of the goitre.

# **Part 2 - Immobilisation of ants after feeding**

Questions to be answered:<br>

1. Does the acidity of the goitre depend on the immobilisation of the ant?<br>
2. Does goitre acidity differ between immobilised ants from different colonies?<br>

```{r}
ant_data_immobilisation <- read.csv("data/Supplementary_Data_Fig.1b.txt", sep="\t")
```

## **EDA**

```{r}
summary(ant_data_immobilisation)
```

```{r}
describe(ant_data_immobilisation)
```

```{r}
str(ant_data_immobilisation)
```

The dataset includes measurements of pH in the goitre of _Camponotus floridanus_ ants across different treatments and colonies. Here are some key observations:

**Treatment**: The treatment column, labeled as `treatment`, likely indicates whether ants were immobilized (FA+ suggests presence of formic acid immobilization treatment, if other categories exist we'll check them).<br>
**Columns**: Includes `species`, `worker` ID, `treatment`, `pH`, and `colony.`

Let's first verify the categories under `treatment` to understand the groups we are comparing, then visualize and analyze the data accordingly.

```{r}
unique(ant_data_immobilisation['treatment'])
```
The `treatment` column has two categories:<br>

**FA+**: Ants that were immobilized.<br>
**FA-**: Ants that were not immobilized.<br>

## **Visualisation**

Let's start by visualizing the pH distribution for immobilized vs. non-immobilized ants and then proceed with the statistical analysis.<br>

```{r}
ggplot(ant_data_immobilisation, aes(x = ph, fill = treatment)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.1, alpha = 0.6, position = "identity") +
  geom_density(alpha = 0.5) +
  facet_wrap(~treatment) +
  labs(title = "Density and Distribution of pH by Treatment")
```

```{r}
ggplot(ant_data_immobilisation, aes(x = treatment, y = ph)) +
  geom_boxplot() +
  labs(title = "Boxplot of pH Values by Treatment")
```

### **Visualization Results**
**Density Plot**: Shows distinct distributions of pH values for immobilized (FA+) and non-immobilized (FA-) ants, suggesting differences in acidity levels depending on the treatment.<br>
**Boxplot**: Highlights the central tendency and variability of pH values for each treatment group. The plot indicates that the pH may be slightly higher on average in non-immobilized ants.<br>
**Data distribution**: Non-normal.<br>

## **Normality test**

Let's use `shapiro.test` to check the distribution of pH values.

```{r}
shapiro.test(ant_data_immobilisation$ph)
```

### **Normality test results**

The data is distributed non-normal

## **Statistical Analysis**
To substantiate the observations from the visualizations, we'll perform:

**Mann–Whitney U test** for independent samples to test if the mean pH values significantly differ between the FA+ and FA- groups.<br>
**Reasons of choice**: Non-normal distributed data. Two independent samples.<br>
**Kruskal-Wallis rank sum test** to assess pH differences between colonies within immobilized ants (FA+).<br>
**Reasons of choice**: Non-normal distributed data. Data with independent observations.<br>

```{r}
fa_plus <- filter(ant_data_immobilisation, treatment == "FA+")
fa_minus <- filter(ant_data_immobilisation, treatment == "FA-")
```

```{r}
wilcox.test(fa_plus$ph, fa_minus$ph, exact = FALSE)
```

```{r}
kruskal.test(ph ~ colony, data = fa_plus)
```

### **Statistical Analysis Results**

**Mann–Whitney U test between FA+ (immobilized) and FA- (non-immobilized)**:
The result shows a highly significant difference in pH values between immobilized and non-immobilized ants ( _p_ = 1.035e-07 ), suggesting lower pH in immobilized ants.
**Kruskal-Wallis rank sum test for pH differences among colonies (FA+ only)**:
The results indicate no significant differences in pH levels among different colonies for immobilized ants ( _p_ = 0.1473), suggesting that colony variation does not significantly impact pH within the immobilized group.

# **Part 3 - Do they pass?**

Question to be answered:<br>
1. Do _Asaia_ and _Serratia_ bacteria pass from the goitre to the midgut?

```{r}
ant_data_pass <- read.csv("data/Supplementary_Data_Fig.5ab.txt", sep = "\t")
```

## **EDA**

```{r}
summary(ant_data_pass)
```

```{r}
describe(ant_data_pass)
```

```{r}
str(ant_data_pass)
```

The dataset consists of 653 entries with the following structure:<br>

**bacteria**: Name of the bacteria (e.g., Asaia, Serratia).<br>
**gutregion**: Part of the gut where the sample was taken (e.g., crop, midgut).<br>
**time**: Timepoint of the sample collection in hours.<br>
**cfu**: Colony-forming units counted.<br>
**colony**: Colony identifier.<br>

## **Visualisation**

Let's start with visualizing the CFU distribution in two different gut regions.

```{r}
ggplot(ant_data_pass, aes(x = cfu)) +
  geom_histogram(bins = 30, fill = "skyblue", color = "black") +
  facet_grid(bacteria ~ gutregion, scales = "free_y") +
  theme_minimal() +
  labs(title = "Distribution of CFU Counts",
       x = "CFU",
       y = "Frequency") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### **Histogram Analysis**
Here are the histograms showing the distribution of colony-forming units (CFU) for both Asaia and Serratia bacteria in different gut regions:<br>

**Asaia in Crop**: The distribution shows a broad range of values with some peaks, indicating variability in the colony counts at different sampling times.<br>
**Asaia in Midgut**: The counts are significantly lower, mostly near zero, suggesting fewer bacteria pass into or survive in the midgut.<br>
**Serratia in Crop**: Similar to Asaia, Serratia shows a varied distribution in the crop but generally at lower counts compared to Asaia.<br>
**Serratia in Midgut**: The counts are very low, similar to Asaia, indicating minimal presence in the midgut.<br>

```{r, warning=FALSE}
ggplot(ant_data_pass, aes(x = interaction(time, gutregion), y = cfu, fill = gutregion)) +
  geom_boxplot() +
  facet_wrap(~bacteria, scales = "free_x") +
  scale_y_log10() +
  labs(title = "CFU Distribution by Bacteria Type Across Time and Gut Regions",
       x = "Time and Gut Region",
       y = "Colony Forming Units (CFU, log scale)",
       fill = "Gut Region") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.background = element_blank(),
        strip.text.x = element_text(size = 12, face = "bold"))
```

### **Boxplot Analysis**
The boxplots display the distribution of colony-forming units (CFU) for both Asaia and Serratia bacteria in different gut regions across various times, with the y-axis on a logarithmic scale for better visualization of the spread of data:<br>

**Asaia in Crop over Time**: There is a noticeable decrease in the median and range of CFU over time, suggesting a decline in bacterial population in the crop as time progresses.<br>
**Asaia in Midgut over Time**: The CFU counts remain consistently low across all time points, with some slight increases but still considerably lower than in the crop.<br>
**Serratia in Crop over Time**: Similar to Asaia, there's a clear decrease in CFU over time, indicating a reduction in bacterial presence in the crop.<br>
**Serratia in Midgut over Time**: The counts are generally low, comparable to those of Asaia, and remain low over time, though there are some outliers or variations at specific times.<br>

## **Normality test**

Let's use `shapiro.test` to check the distribution of CFU values.

```{r}
ant_data_pass %>%
  group_by(gutregion, time) %>%
  shapiro_test(cfu)
```

### **Normality test results**

The data is distributed non-normal

## **Statistical Analysis**

Statistics will be done using Kruskal-Wallis Test, because:<br>
**Non-Parametric Test**: Kruskal-Wallis Test does not require the data to follow a normal distribution. This is particularly useful since the distribution of CFU counts, as seen in the histograms, does not clearly follow a normal distribution.<br>
**Comparative Analysis**: Our goal is to compare bacterial counts between two samples (crop vs. midgut) at various time points.<br>

```{r}
ant_data_pass %>% 
  group_by(bacteria, time) %>% 
  kruskal_test(cfu ~ gutregion)
```

Now we perform post hoc test with Dunn's test

```{r}
ant_data_pass_results_posthoc <- ant_data_pass %>%
  group_by(bacteria, time) %>%
  dunn_test(cfu ~ gutregion, p.adjust.method = "bonferroni")
ant_data_results_posthoc
```

### **Statistical Analysis Results**

Before interpreting the results, it should be remembered that for the first four hours the food remains in the goitre and only a small part of it goes to the midgut. Only after 18 hours the bulk of the food passes from the goitre to the intestine (midgut).<br>

For _Asaia_, there is a statistically significant difference between the crop and midgut CFU counts at all measured time points. This suggests that there is a consistent difference in the bacterial population between these two gut regions across time. Looking at the boxplot, it can be seen that the CFU count of _Asaia_ becomes higher after 18 hours (24 and 48 hours). At 24 hours, when comparing the GI tract sections using Kruskal Wallis test, the statistic is -2.9663418 and the p-value is 3.013653e-03, which indicates that 24 hours after ingestion of food, the _Asaia_ bacteria have passed into the midgut, but are in lower numbers than in the goitre. But at the time interval of 48 hours, the value of the statistic is already 3.2963553 with an extreme low p-value of 9.794805e-04, which means that 48 hours after ingestion of food, the number of _Asaia_ bacteria in the midgut is higher than in the goitre.<br>

For _Serratia_, significant differences are observed in the early time points (0, 0.5 and 4 hours), which means absolutely nothing based on the fact that bacterias are expected to come up to midgut after 18 hours. Especially these resuts are of no use to us since _Seratia_ bacteria has 0 CFU in midgut at 4 hours! However, at 24 and 48 hours, the p-values indicate no significant differences, which might suggest that over time, any initial differences in population may equalize. Yet in boxplot it can be seen that at 48 hours there are 0 _Seratia_ CFUs in midgut. So only 24 hours results are interesting for us.

**Summary:** Bacteria from both of these genera pass into the intestine. However, bacteria from the genus _Asaia_ stay there for a long time, while bacteria from the genus _Seratia_ do not stay there longer than 24 hours.

# **References**

The following packages must be installed to work with the dataset:

```{r}
knitr::write_bib(c("ggplot2", "dplyr", "stats", "psych",
                   "readr", "tidyr", "ggpubr",
                   "purrr", "rstatix"), file = "packages.bib")
```